Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-20 main.tex
in IX again

\todo{add $\cfaD$ everywhere}

In this section we establish the relation between skew bifibrations and derivations  in
$\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, shown below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
\todo{same for $\cfaD$}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

Note that in an instance of $\rectif\mfaD$ or $\rectif\mexD$ (as shown
above), we can have $x=y$ or $x=z$, but not both if the premise is
rectified. If $x=y$ and $x=z$ we have $\mfaD$ and
$\mexD$ as special cases of $\rectif\mfaD$ and $\rectif\mexD$, respectively.

@@

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr=\wrD$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (as acts as the identity on all other vertices). For a
derivation $\Deri$ in
$\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$. \juihsuan{maybe mention at least that the induced
maps define graph homomorphisms. Do we need to talk about the contexts
$\Sconhole$ here (induced maps act clearly as the identity on contexts but we need them for the composition)?}

\begin{lemma}
  Let $\downsmash{\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
      B}{ \vlhy{\rectif A}}}$, such that the induced maps
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
    B}$ are equal up to the relabeling of the vertices.
\end{lemma}

@@

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wrD,\acD,\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a rectified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

>>>>

In this section we establish the relation between skew bifibrations and derivations  in
$\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\cfaD$, $\mfaD$, and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\cfaD$,  $\rectif\mfaD$ and
$\rectif\mexD$, shown below:
\begin{equation*}
  \vlinf{\rectif\cfaD}{}{\Scons{\forall x.Ax}}{\Scons{\forall y.\forall x.Ax}}
  \hskip3em
  \begin{array}{c}
    \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
    \\ \\[-1ex]
    \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
  \end{array}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. We also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

Note that in an instance of $\rectif\mfaD$ or $\rectif\mexD$ (as shown
above), we can have $x=y$ or $x=z$, but not both if the premise is
rectified. If $x=y$ and $x=z$ we have $\mfaD$ and
$\mexD$ as special cases of $\rectif\mfaD$ and $\rectif\mexD$, respectively. And similarly,  if $x=y$ then $\cfaD$
is a special case of $\rectif\cfaD$.

@@

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr=\wrD$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\cfaD,\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (as acts as the identity on all other vertices).
For a
derivation $\Deri$ in
$\set{\wrD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.
\juihsuan{maybe mention at least that the induced
maps define graph homomorphisms. Do we need to talk about the contexts
$\Sconhole$ here (induced maps act clearly as the identity on contexts but we need them for the composition)?}

\begin{lemma}
  Let $\downsmash{\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\rectif\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
      B}{ \vlhy{\rectif A}}}$, such that the induced maps
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
    B}$ are equal up to a variable renaming of the vertex labels.
\end{lemma}

@@

We say that a derivation $\Deri$ is \bfit{sane} if for every line $Q$
in $\Deri$ we have that $\graphof D$ is a fograph (i.e., all binders
are legal). Clearly, every rectified derivation is sane, but not vice
versa, as we might have multiple occurrences of bound varibles in $Q$,
such that $\graphof D$ is still a fograph.

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wrD,\acD,\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a sane derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}
